Inferring causal connectivity from pairwise recordings and optogenetics

### 2018-06-25 manuscript/main.tex
updates

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx,subcaption}
\usepackage{todonotes} %[disable]
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usetikzlibrary{arrows,shapes,calc}
\tikzset{line/.style = {draw, -latex'}}
\newcommand{\abs}[1]{\left|{#1}\right|}
\newcommand{\pde}[2]{{\frac{\partial {#1}}{\partial {#2}}}}
\newcommand{\pdde}[2]{\frac{\partial^2 {#1}}{\partial {#2}^2}}
\newcommand{\de}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\dde}[2]{\frac{\mathrm{d}^2 {#1}}{\mathrm{d} {#2}^2}}
\newcommand{\intdef}[4]{\int\limits_{#1}^{#2} #3\, \mathrm{d}{#4}}
\crefname{appendix}{Appendix}{appendices}
\crefname{section}{Section}{sections}
\crefname{table}{Table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{algorithm}{Algorithm}{algorithms}
\Crefname{algorithm}{Algorithm}{Algorithms}
\crefname{equation}{Eq.}{eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Fig.}{figs.}
\Crefname{figure}{Figure}{Figures}
\crefname{subfigure}{Fig.}{figs.}
\Crefname{subfigure}{Figure}{figures}
\captionsetup[subfigure]{labelformat=simple}
\renewcommand{\thesubfigure}{(\alph{subfigure})}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
%opening
\title{Correctly inferring causal connectivity using optogenetics}
\author{Mikkel Elle Lepper\o{}d, Konrad Paul Kording}

@@

\begin{abstract}
\noindent Neurons interact through spikes and a central objective of neuroscience is measuring the way how neurons causally affect one another. To probe such interactions, scientists often use optogenetics which typically leads to stimulation effects on multiple cells. This then produces a so-called confounding problem - we cannot know which of the stimulated neurons affected the activity of a given target neuron. Here we show how the resulting biases are large and how causal inference techniques, in particular instrumental variables, from econometrics can ameliorate this problem. Our approach utilizes the fact that neurons have absolute refractory times where the stimulation has no influence, and the missing stimulation acts like an effectively random variable which can then be used to disentangle causality. If we can not apply truly random stimulation to a network of neurons then causal inference techniques are needed to identify mechanisms.
\end{abstract}

\section{Introduction}
% neuroscience experiments are observational which implicates a certain set of statistical tools
% spike timing cannot give a causal estimate of connectivity, thus we need to perturbe the system
%a quest for causality and optogenetics is a main tool

%set up the drive to causality
We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.

%we generally use observational studies
We use observational studies. But causality is typically wrong. <nutshell version of the mehler paper>

%set up confounding
 The experiments are extremely hard to control, even with the most accurate measurements of environmental variables and minimal genetic variance. Actually, given the vast amount of neurons and their connections in primitive mammals experiments are to be considered an observational sort. In a quest for experimental control we are now able to measure from thousands of neurons at the same time with micro-second precision by electrophysiology or with high spatial accuracy by two photon imaging. However, the large amount of unmeasured activity makes estimates of functional connectivity confounded. By confounded we mean that correlations between neuron pairs are largely influenced by the background activity. One example of this confounding is when two neurons $ A,B $ are driven by a common input $ D $, while one wants to estimate connectivity between $ A $ and $ C $. However only $ B $ and C is connected such that B and C are strongly correlated, then since $ B $ and $ C $ is correlated then $ A $ and $ C $ is correlated and a regression $ C = \beta A $ will give an erroneous $ \beta > 0 $ due to the the $ A,B $ correlation given by their common input $ D $. In this case we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. To estimate causal relationships between neurons, stimulation is appropriate, and if able to stimulate single neurons the ability to estimate causal relationships by regression is within grasp. However, this is experimental challenging and yield very low cell count. Therefore an optogenetic approach is more than often preferable.

%optogenetics is totally nonlocal
Within the bounds of reachable tissue for photons emitted by a light source the decay of light intensity decays proportional to $ 1/r^2 $. Furthermore if the distribution of neurons is homogeneous the probability of affecting a neuron increases in distance by $ r^2 $. This essentially makes the distribution of affected neurons flat. When performing simultaneously recording with optogenetic stimulation we again get a confounding problem when estimating the functional relationships between neuron pairs. In this case the confounding is when several neurons see the same stimulus and thus become highly correlated during stimulation yielding statistically indistinguishable functional relationships, this case is similar to the above example where $ D $ now is the optogenetic stimulation. Flat distribution of V.

%causal inference is a thing, if not done well you say stupid things
The problem of endogenous regressors is common in microeconomics and in the social sciences\todo{cite}, where observed outcomes can be driven by a large number of unknown factors. It is thus largely recognized that special care has to be taken when attempting to infer causal relationships between a putative actor and an outcome.

%example
Let's say that you want to estimate the return $ \beta $ from education $ x $ to yearly income $ y $ with the regression $ y = \beta x + u $, where $ u $ is the factors other than schooling that contributes to yearly income. One of the factors in $ u $ is a persons ability, however this may also affect schooling and thus the regressor is correlated with the error term and a regression estimate will not estimate the magnitude of causation from schooling. In this case one may use the proximity to a college or university as an instrumental variable (IV) $ z $\todo{cite card 1995, found in cameron}. There are two criterion that must be fulfilled when choosing an instrumental variable, the instrument must be (1) uncorrelated with the error term, and (2) correlated with the regressor $ x $. Then one may for instance estimate $ \beta $ by a two stage ordinary least squares (OLS) where the first stage is an OLS of the regressor and the instrument which is further used in the second stage with the dependent variable $ y $\todo{cite TSOLS}. summary sentence.

%worry is this really correct? Yes, its provable. 
worry is this really correct? Yes, its . IVs are (under the right assumptions) actually correct.

%here we
The IV technique can also be employed if one seeks to estimate the causal connectivity between neuron pairs. In this study we begin by showing how a simple network of three leaky integrate and fire (LIF) neurons can be heavily confounded under the right circumstances when simulating optogenetic targeting as direct current pulses. We then show that using the refractory period as an instrumental variable we are able to produce good estimates of causal connectivity. Furthermore we show that the method works in a large (N=1250) network of LIF neurons. With this data at hand we compare the IV method with a logistic regression.

\section{Results}
\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{\textwidth} \includegraphics[scale=1]{light-decay}
\caption{} \label{fig:concept:1}
\end{subfigure}

\medskip
\begin{subfigure}{\textwidth} \includegraphics[scale=1]{opto-powerlaw}
\caption{} \label{fig:concept:2}
\end{subfigure}
\caption{Sketch of the problem at hand \label{fig:concept}}
\end{figure}
\todo[inline]{perturbation should spread by 1/r2 and all perturbed connections should be tested, here logit will fail.}
\todo[inline]{figure showing stimulating of network}
\todo[inline]{figure comparing methods with real weights}
\todo[inline]{figure showing power (true false vs n stimulations)}
\todo[inline]{data showing false positive/negative comparing stimulation patterns, comparing stimulation durations}
\todo[inline]{data showing synchrounous regular does not work (can be put together with regular stim. pattern)}
\todo[inline]{Should we show that ccg or gml give lots of false positives? without considering stimulation?}
\todo[inline]{I think I would choose a couple of examples that emulate what an experimentalist would do
(a) periodic slow (bad power not enough data)
(b) periodic fast (good power but confounded)
(c) random fast (good everything)
But I would then also redo all of them with a "weak network" model where therefore the assumptions are pretty well matched an also a "strong network" model where neurons have massive influence on one another (maybe oscillate or something) and hence the weakness assumption that is behind the IV justification breaks down }
Consider three neurons (A,B,C) where A and B receives optogenetic stimulation and only B is connected to C \cref{fig:intro}\labelcref{fig:intro:1}. The stimulation induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:1,fig:intro:2,fig:intro:3,fig:intro:4}, and thus the monosynaptic connection between B and C is thus spuriously inherited between A and C confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC statistically indistinguishable as seen in \cref{fig:intro}\labelcref{fig:intro:5,fig:intro:6} respectively.
\begin{figure} 
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1.3]{simple}
\caption{} \label{fig:intro:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{traces}
\caption{} \label{fig:intro:2}
\end{subfigure}

\medskip
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{raster}
\caption{} \label{fig:intro:3}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr}
\caption{} \label{fig:intro:4}
\end{subfigure}
\caption{Sketch of a simple network containing three neurons depicted in red stimulated with blue laser light trough optogenetic activation. Invoked spikes due to stimulation is depicted in (b) upper panel with corresponding voltage traces in the lower panel. The neurons A and B are stimulated multiple trials and the  histogram of all trials are shown in (c) upper panel with the corresponding raster plot for each trial in the lower panel. Horizontal and vertical axes in correlograms represents respectively time lag in ms and counts of coincident spikes in bins of 1 ms for \labelcref{fig:intro:4} \label{fig:intro}}
\end{figure}

\begin{figure}
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr_highres_BC}
\caption{} \label{fig:cchvswald:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr_highres_AC}
\caption{} \label{fig:cchvswald:2}
\end{subfigure}

\medskip
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{wald_BC}
\caption{} \label{fig:cchvswald:3}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{wald_AC}
\caption{} \label{fig:cchvswald:4}
\end{subfigure}
\caption{Cross correlograms for all neurons are depicted in (d) and the high resolution zoom of BC and AC are visualized in (e) and (f) respectively. Horizontal and vertical axes in correlograms represents respectively time lag in ms and counts of coincident spikes in bins of 0.1 ms for \labelcref{fig:cchvswald:1,fig:cchvswald:2}. \label{fig:cchvswald}}
\end{figure}

\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.2]{network_stim}
\caption{} \label{fig:network-power:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth}\includegraphics[scale=.2]{wald_network}
\caption{} \label{fig:network-power:2}
\end{subfigure}

\caption{Power analysis of IV estimator. \label{fig:network-power}}
\end{figure}

\begin{figure}
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{false_positive}
\caption{} \label{fig:network-class:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{false_negative}
\caption{} \label{fig:network-class:2}
\end{subfigure}

\medskip
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{fit_wald}
\caption{} \label{fig:network-class:3}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{fit_cch}
\caption{} \label{fig:network-class:4}
\end{subfigure}
\caption{Comparing CCH and logit with IV \labelcref{fig:network-class:1,fig:network-class:2}. \label{fig:network-class}}
\end{figure}

\section{Discussion}
Spike time methods works well when network is desynchronized, small amplitude stimulation with close range. Our method opens up for the possibility of strong perturbations giving larger yield.
We are making the network results quite hard as all synapses have the same time constant such that multiple neurons are affecting the same cell at the same time for each stimulation

%Correlated refractory periods is *the* failure mode. How to randomize them

% Maybe nonlinearities in stimulation rescue old school approaches
Discuss distribution of V. Discuss 2p stimulation.  Conclude that there is absolutely no reason to hope things will work with 1p.

%Maybe some biophysics pecularity will rescue us
No way. But cite Buszaki.

%We recommend the WALD 
easily usable in the context of many experiment. Potentials for future improvement.

%brain regions versus neurons
IVs on multiple scales.

%Causal inference techniques to fix experimental issues
Broad opportunity, cite RDD, cite diff-in-diff etc, broad overview of the promise of econometrics for neuro, matching

%Constructing good IVs
Multiple illuminations, designed sticky opsins

\FloatBarrier
\section{Methods}
\subsection{Instrumental variable estimation}
A simple approximation of the connectivity strength between putative sender and receiver neurons $ x $ and $ y $ respectively can be to ignore long distance excitation and simply calculate the relation between the spike times in $ x $ and $ y $ with a regression model given by

\begin{equation}
y = \beta x + u.
\label{eq:regress}
\end{equation}
Here $ y $ is the dependent variable, $ x $ is the explanatory variable, $ \beta $ is the slope of the $ x,y $ curve and $ u $ is an unknown error term. This system follows the causal path:
\begin{center}
\begin{tikzpicture}
    \node (u) {$ u $};
    \node [above = of u] (x) {$ x $};
    \node [above right = of u] (y) {$ y $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (y);
\end{tikzpicture}
\end{center}
Assuming that changes in spike times $ y $ are described by $ \beta x $ i.e. $ \de{y}{x} = \beta $ for spike times $ x $. One problem with this idea is that in a confounded system, perfectly correlated neurons will give statistically indistinguishable $ \beta $. In the extreme case where two neurons are both made to fire every time they are stimulated, they will have the same weights according to \cref{eq:regress}, after all, during stimulation  $ y=1 $ for both, even if only one of them drives the downstream neuron. Another problem is if the network state affects both the probability of a neuron to fire and also the probability of downstream neurons to fire. In this case, the network state can induce a correlation which will make the estimation highly biased. Arguably, the network state will, in all realistic models, have a dramatic influence on all neurons. In general, if multiple neurons are stimulated synchronously our estimates will be off, potentially massively so as seen in the equation
\begin{equation}
y = \beta x + u(x).
\end{equation}
Following the causal path

\begin{center}
\begin{tikzpicture}
    \node (u) {$ u $};
    \node [above = of u] (x) {$ x $};
    \node [above right = of u] (y) {$ y $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
\end{tikzpicture}
\end{center}
We now have the relation $ \de{y}{x} = \beta + \de{u}{x} $. To get at causality we thus require some stimulation that only highlight the activity in $ y $ caused by $ x $, however optogenetic stimulation is not specific enough as the photons will activate parts of the network activity $ u $. To have a remedy, we need something that can distinguish between different neurons that are stimulated. We thus require some instrument $ s_{r} $ which is (1) uncorrelated with the network $ u $ and (2) is correlated with the regressor $ x $\todo{cite the book or whatever}. The graph with stimulation is illustrated by

\begin{center}
\begin{tikzpicture}
    \node (u) {$ u $};
    \node [above = of u] (x) {$ x $};
    \node [above right = of u] (y) {$ y $};
    \node [left = of x] (s) {$ s $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
    \path [line] (s) -- (x);
    \path [line] (s) -- (u);
\end{tikzpicture}
\end{center}
The previous graph represents a confounded stimulation, however we may use the fact that a neuron that has fired just before the stimulation will be in absolute refractory period and hence have $ x_{s}=0 $. This introduces times where the spike from one of the stimulated neurons is missing. Neurons have low firing probability once we use small bins and are very noisy at spiking at small time scales. Thus if we assume that (1) the stimulation pattern is random (2) the network activity $ u $ is asynchronous and irregular we may use the refractory period as an instrumental variable illustrated in the following graph

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{breakcites}
\usepackage{graphicx,subcaption}
\usepackage{todonotes} %[disable]
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usepackage{booktabs}
\usepackage{natbib}
\usetikzlibrary{arrows,shapes,calc}
\tikzset{line/.style = {draw, -latex', ultra thick}}
\newcommand{\abs}[1]{\left|{#1}\right|}
\newcommand{\pde}[2]{{\frac{\partial {#1}}{\partial {#2}}}}
\newcommand{\pdde}[2]{\frac{\partial^2 {#1}}{\partial {#2}^2}}
\newcommand{\de}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\dde}[2]{\frac{\mathrm{d}^2 {#1}}{\mathrm{d} {#2}^2}}
\newcommand{\intdef}[4]{\int\limits_{#1}^{#2} #3\, \mathrm{d}{#4}}
\newcommand{\mean}[1]{\left<{#1}\right>}
\crefname{appendix}{Appendix}{appendices}
\crefname{section}{Section}{sections}
\crefname{table}{Table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{equation}{Eq.}{eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Fig.}{figs.}
\Crefname{figure}{Figure}{Figures}
\crefname{subfigure}{Fig.}{figs.}
\Crefname{subfigure}{Figure}{figures}
\captionsetup[subfigure]{labelformat=simple}
\renewcommand{\thesubfigure}{(\alph{subfigure})}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
%opening
\title{Inferring causal connectivity from pairwise recordings and optogenetics}
%\title{The need to combine optogenetics with neural recordings to estimate causal connectivity}
\author{Mikkel Elle Lepper\o{}d, Konrad Paul Kording}

@@

\begin{abstract}
\noindent Neuroscientists often want to identify how neurons causally interact. Perturbations with optogenetics are a popular way of quantifying such interactions. However, even focused light will typically stimulate multiple cells, which produces a confound - we cannot know which of the stimulated neurons affected the activity of a given target neuron. Here we show how this confound produces large biases in estimates of connectivity, and how we can reduce these biases by a data analysis strategy that exploits the absolute refractory period of neurons. We use the instrumental variable (IV) framework from economics which estimates causal strengths by exploiting a variable that affects only one of the variables, here the presynaptic neuron. The interaction between optogenetic stimulation and the absolute refractory period produces a weak almost random signal (IV) that we exploit to solve the causal inference problem. In a simple simulation of integrate and fire neurons, we find that instrumental variable techniques estimate connectivity better than naive techniques (99.8\% vs 86\% correct). Our instrumental variable approach is robust and easy to implement in the context of current experimental paradigms.
\end{abstract}

\section{Introduction}
% neuroscience experiments are observational which implicates a certain set of statistical tools
% spike timing cannot give a causal estimate of connectivity, thus we need to perturbe the system
%a quest for causality and optogenetics is a main tool
% We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.
%set up the drive to causality
We want to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, action, and cognition. For such an understanding it is not sufficient to know the correlations between variables or even be able to predict them. After all, there can be many ways how the same activities come about by distinct causal chains \citep{drton2011global, peters2017elements}. Identifiability of causal networks is complicated when it comes to brain data. Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. Thus, brain data which is almost always observational, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}. While observing correlations within the system is usually relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard or, more commonly, impossible.  Getting at such an understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons. Only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}. Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.

% we use observational studies. But causality is typically wrong. 
Moreover, we generally only record from a small subset of all neurons. The data we obtain from typical recordings, e.g. from electrophysiology  or calcium imaging, is very low dimensional relative to the dimensionality of the brain. Moreover, it is observational, which means that it does not result from randomized perturbations. In such cases, we can never know to which level the observed activity was caused by other observed activity, or by the activity of the unobserved neurons. Such unobserved activity is then called confounders. If mechanisms are estimated from observational data in the presence of confounders, we will generally make large errors and draw incorrect conclusions \citep{angrist2008mostly}. Unobserved neural activity confounds estimates of causal interactions and makes it difficult to estimate underlying mechanisms.

%set up confounding
Confounding is the big threat to causal validity \citep{UpcomingMehlerPaper} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.  A much used method for estimating the output of single neurons is to perform multiple regression analyses \citep{pillow2008spatio}, modeling each neuron with a generalized linear model (GLM).  Multiple regression may be perceived as a solution to confounding problems as they support the concept of ``explaining away" \citep{stevenson2008inferring}. However, in current work we will exclude approaches that look at multiple neurons at the same time. This is because, ``explaining away" can only be a good strategy if most neurons are included in the recordings but this is rarely the case for most experimental settings, especially in the mammalian brain. We will therefore focus on neuron pairs where stimulus and the pre- and post-synaptic neurons are known.

To estimate connectivity it is first and foremost important that it stems from cause and effect e.g. B's action potential is causing an increase in C's membrane potential, therefore we use the term causal connectivity.  For a given example we want to estimate causal connectivity between two neurons, $ A $ and $ C $ (\cref{fig:intro}\labelcref{fig:intro:1}). Two neurons, $ A $ and $ B $, are driven by a common input $ D $ and because $ B $ and $ C $ are connected they are strongly correlated. Consequently, $ A $ and $ C $ are also correlated and a regression  $ C = \beta A $ will misleadingly conclude that there is a direct interaction if it was causally interpreted.  In this case we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. Na\"ive regressions in partially observed systems will generally not reveal causality.

To estimate causal relationships between neurons, stimulating the pre-synaptic neuron is the gold standard. In fact, a common definition of causality is in terms of what would happen if one would change the value of one variable in the system, independently of changing other variables -- an intervention \citep{pearl2009causality}. If we stimulate single neurons, the ability to estimate causal relationships by regression is within reach. However, this is experimentally challenging and yield low cell count because it either requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}. Because gold-standard perturbations are challenging, it would be highly desirable if causality could be obtained from optogenetic stimulation  \citep{boyden2005millisecond, zemelman2002selective}.

%optogenetics is totally nonlocal
Interpreting the results of optogenetic stimulation in terms of causal interactions is difficult. The expression of opsins are often restricted to a population of neurons, but regular optogenetic stimulation will still affect many neurons simultaneously. In \textit{in vivo} settings it is generally impossible to direct photons to only one neuron. Hence, the stimulus will produce a distributed pattern of activity. For example, if we focus a stimulation beam on one neuron, there will be a cone of light in front of and behind the neuron, which comes from the rays of light from the lens focused on the cell and again coming out on the backside.. This distributed pattern of stimulation produces activity which then percolates through the network of neurons. Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons introducing problematic confounders.

%causal inference is a thing, if not done well you say stupid things
The inference of causality from observational data is addressed in the fields of statistics \citep{pearl2009causality}, machine learning \citep{peters2017elements} and econometrics \citep{angrist2008mostly}. Within these fields, the problem of endogenous regressors is commonly addressed. We may thus look towards these fields for insights into how we may resolve the confounding problem induced by optogenetic stimulation.

%example
A commonly used approach towards causal inference in economics are instrumental variables. Let's say that we want to estimate the return $ \beta $ from education $ x $ to yearly wages $ y $ with the regression $ y = \beta x + u $. Here $ u $ are the factors other than education that contribute to yearly wages. One of the factors in $ u $ is a person's cognitive ability. However, a person's cognitive ability may also affect education and thus the regressor $ x $  is correlated with the error term $ u $. This will imply that the regression estimate $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association. In this case one may use the proximity to a college or university as an instrumental variable (IV) \citep{card1993using}. Proximity is a good instrument since it can only affect wages through schooling, since we expect living in proximity of colleges and universities give higher probability to attend but not to affect other contributing factors to wages such as cognitive ability. Then, in order to attribute the causal effect of education on wages one may calculate the ratio of covariances $ \beta = cov(\mathrm{proximity},\mathrm{wages})/cov(\mathrm{proximity},\mathrm{education})$. This ratio corrects for the confounding factor.

%worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
The IV technique has been used extensively in econometrics and is provable causal given three assumptions \citep{angrist2008mostly}. First, the instrument must be uncorrelated with the error term. Second, the instrument must be correlated with the regressor. Third, there must be no direct influence of the instrument on the outcome variable but only an influence through the regressor variable. The validity of these assumptions is central when using the IV.

% Refractory period as IV
For an instrument to be good, it needs to be unaffected by other variables. In the brain, almost everything is affected by the network state. However, certain variables can be more or less affected. For example, the overall activity of the network is through slow and strongly nonrandom dynamics. In contrast, the temporal pattern of when a neuron is in a refractory state may be random. First, if neurons are spiking according to conditional Poisson distributions, their exact timing will, conditioned on the network state, be random. While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is likely quite random.

%here we
Here we show that the instrumental variable (IV) technique can be employed if one seeks to estimate the causal connectivity between neuron pairs. We begin by showing how confounding factors are introduced by regular optogenetic stimulations. We then simulate this confounding effect in a simple network of three leaky integrate and fire (LIF) neurons. With this simple model we show that by using the refractory period as an IV we are able to distinguish between connected and unconnected neuron pairs. We compare these estimates with a naive although widely used cross-correlation histogram (CCH) method that fails to distinguish respective pairs. We then turn to a simulated network of randomly recurrent connections of excitatory and inhibitory LIF neurons with distributed synaptic weights. With this data at hand we first calculate the mean squared errors of the IV method and show that it is robust to different simulated network states. Finally, we compare the amount and size of false positive and false negative estimates and goodness of fit on synaptic weights with pairwise assessments using CCH and logistic regression.

\section{Results}
\subsection{Optogenetics is not local}
Optogenetics is generally seen as a perturbation method that by and large affects neurons in close proximity of the light source. However, it can be questioned if this is the correct way of conceptualizing the spatial effect of stimulation. The stimulation effects depend on multiple factors. Firstly, light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. Secondly, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity. Lastly, physiological properties of the cells are important as light may e.g. only have a strong effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold. The induced effect of optogenetic stimulation as a function of distance should be the product of four parameters:  light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and opsin distribution across neurons.

To estimate the light intensity we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \cite{Aravanis2007}; see \cref{sec:method:opto}. This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). This is explained by the surface of the 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic beyond the scattering length \cref{fig:concept} (inset). The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed. The density of photons thus decreases rapidly with distance.

The number of illuminated neurons at a given distance will be larger the further away they are from the stimulation site given that neurons are uniformly distributed in brain tissue \cref{fig:concept} (black line). In fact, it will increase by approximately $ r^2 $ with distance. This derives from the same surface scaling as for the 3D shell. Thus the number of neurons that can be activated increases rapidly with distance.

To estimate the effect of stimulation, the functionality underlying spiking activity needs to be considered. This can largely be characterized by the  distribution of membrane potentials across neurons. Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}. The response that is expected due to a pulse of light that induces a charge $Q$ should be proportional to the density of neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance). The fact that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the response in membrane potential to a perturbation for any neuron is roughly proportional to the light intensity. Moreover, assuming that opsins are evenly distributed along the neuron, the opsin density in the neuron population will be relatively even. Based on this, we calculate the overall stimulation effect to be the product of neuron density and light intensity which is approximately constant in distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). Thus, optogenetic stimulation utilizing single photon activation does not actually produce a localized effect.

\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
\caption{{\bf Spatial extent of optogenetic stimulus}. Due to scattering and geometric loss the light intensity (I, cyan line) plotted as percentage of intensity exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber. If neurons are uniformly distributed the number of affected neurons increase by $ r^{2} $ (N, black line) rendering the probability of activating a neuron approximately constant (IN, blue line). \label{fig:concept}}
\end{figure}

\subsection{Confounding as a problem for the estimation of causal effects}
When we stimulate many neurons at the same time, and observe a post-synaptic neuron to be active after our stimulation, it is hard to know which of the stimulated neurons produced the activity. To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C). The neurons receive Poisson spike trains and have added Gaussian white noise to the membrane potential. We also gave them interactions, where spikes of neuron B increase the probability of firing for neuron C but there are no other interactions \cref{fig:intro}\labelcref{fig:intro:1}.
Finally, we allowed simulated optogenetic stimulation to affect neurons A and B (but not C). We thus have a simple system for exploring questions of causality.

After running the simulation, the peri-stimulus time histogram of the stimulated neurons show the result of both the stimulation itself (suppressed for visibility) and the neuron's refractory period \cref{fig:intro}\labelcref{fig:intro:2} (AA, BB).  Since the stimulation affects A and B simultaneously, it induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:3} (AB). This further generates a strong correlation between A and C, confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}) shown in \cref{fig:intro}\labelcref{fig:intro:3}. Even though the correlation peak between B and C is larger than between A and C due to correlated spikes outside the periods with stimulation one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected. If stimulation affects multiple neurons simultaneously, there is a real confounding problem.

\begin{figure} 
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{simple}
\caption{} \label{fig:intro:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{raster}
\caption{} \label{fig:intro:2}
\end{subfigure}

\medskip
\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
\caption{} \label{fig:intro:3}
\end{subfigure}
\caption{{\bf A,B,C network}. A sketch of the simple network containing three neurons %depicted in \labelcref{fig:intro:3} 
shows stimulation configuration with blue laser light and the connections with arrows (a). The neurons A and B are stimulated 1000 trials and the corresponding peristimulus time histogram are shown in \labelcref{fig:intro:2} upper panel with a raster plot in the lower panel. Cross correlation histograms (CCHs) are shown in \labelcref{fig:intro:3} where horizontal and vertical axes represents time lag in ms and counts of coincident spikes in bins of 1 ms. \label{fig:intro}}
\end{figure}

\subsection{Instrumental variables to resolve confounding}
In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons we need something that can distinguish the influence of one stimulated neuron from the influence of another stimulated neuron. We would thus need something that affects the stimulation effect separately between neurons. Arguably, refractoriness is such a variable. If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike. This gives us an interesting way of getting at causality, by comparing the network state between a time when a neuron is able to spike and a time where the neuron is unable to spike.

Instrumental variables requires the existence of a random (or sufficiently random) variable that affects a variable of interest. This independent influence among neurons then allows quantifying the influence of the variable of interest on the rest of the network. In our case, the refractory states of a neuron is in good approximation independent on small time scales (see Discussion for caveats). It affects the influence of stimulation on the potentially pre-synaptic neuron (\cref{fig:cchvswald}\labelcref{fig:cchvswald:0}). The trials where a stimuli is unable to elicit a spike due to the refractory state can then be used to identify causal effect on the potentially post-synaptic neuron.

We can now investigate if the use of an instrumental variable gives a better estimate of connectivity strength than simply analyzing the lagged correlations by means of the CCH; see \cref{eq:ptrans}. We thus use the IV estimator \cref{eq:wald} on the three neuron system (\cref{fig:cchvswald}\labelcref{fig:cchvswald:1,fig:cchvswald:2}). It converges to the correct causal conclusions that the weights $ w_{BC} = 0.2 $ and $ w_{AC} = 0 $. For such a simple system, it produces meaningful estimates of the causal interactions between neurons.
